\name: Deterministic and Coherent Implicits

### 2018-05-17 jfp/src/Translation.tex
address various small issued raised by review 3
additional lemmas and theorems

%-------------------------------------------------------------------------------
\subsection{Type-Directed Translation}
Figure~\ref{fig:type} presents the translation rules that convert $\name$
expressions into ones of System F. 
The gray parts of the figure extend the type system with the necessary
information for the translation.

@@

Moreover, we can express three key properties of Figure~\ref{fig:resolution2}'s
definition of resolution in terms of the generated evidence.
Firstly, the deterministic version of resolution is a sound variation on the original ambiguous resolution.
\begin{lemma}[Soundness]
Figure~\ref{fig:resolution2}'s definition of resolution is sound (but
incomplete) with respect to Figure~\ref{fig:resolution1}'s definition.
\[\forall \tenv, \rulet, E: \quad\quad \dres{\tenv}{\rulet}{E} \quad\Rightarrow\quad \ares{\tenv}{\rulet}{E} \]
\end{lemma}
Secondly, the deterministic resolution guarantees a strong form of coherence:
\begin{lemma}[Determinacy]
The generated evidence of resolution is uniquely determined.
\[\forall \tenv, \rulet, E_1, E_2: \quad\quad 
     \dres{\tenv}{\rulet}{E_1} ~\wedge~ 
     \dres{\tenv}{\rulet}{E_2} \quad\Rightarrow\quad E_1 = E_2 \]
\end{lemma}
Thirdly, on top of the immediate coherence of deterministic resolution, 
an additional coherence property holds.  
\begin{lemma}[Stability]
Resolution is stable under type substitution.
\[\forall \tenv,\alpha,\tenv',\sigma,\rulet, E: \enskip\
\dres{\tenv,\alpha,\tenv'}{\rulet}{E} \enskip\wedge\enskip \tenv \vdash \sigma
\enskip\Rightarrow\enskip 
\dres{\tenv,\tenv'[\sigma/\alpha]}{\rulet[\sigma/\alpha]}{E[\ttrans[\sigma]/\alpha]} \]
\end{lemma}

>>>>

%-------------------------------------------------------------------------------
\subsection{Type-Directed Translation}
Figure~\ref{fig:type} presents the translation rules that convert $\name$
expressions into ones of System~F. 
The gray parts of the figure extend the type system with the necessary
information for the translation.

@@

Moreover, we can express three key properties of Figure~\ref{fig:resolution2}'s
definition of resolution in terms of the generated evidence.
Firstly, the deterministic version of resolution is a sound variation on the original ambiguous resolution.
\begin{theorem}[Soundness]
Figure~\ref{fig:resolution2}'s definition of resolution is sound (but
incomplete) with respect to Figure~\ref{fig:resolution1}'s definition.
\[\forall \tenv, \rulet, E: \quad\quad \dres{\tenv}{\rulet}{E} \quad\Rightarrow\quad \ares{\tenv}{\rulet}{E} \]
\end{theorem}
Secondly, the deterministic resolution guarantees is effectively deterministic:
\begin{theorem}[Determinacy]
The generated evidence of resolution is uniquely determined.
\[\forall \tenv, \rulet, E_1, E_2: \quad\quad 
     \dres{\tenv}{\rulet}{E_1} ~\wedge~ 
     \dres{\tenv}{\rulet}{E_2} \quad\Rightarrow\quad E_1 = E_2 \]
\end{theorem}
It follows immediately that deterministic resolution is also coherent,
which is formulated as follows for our elaboration-based setting.
\begin{corollary}[Coherence]
\[\forall \tenv, \rulet, E_1, E_2: \quad\quad 
     \dres{\tenv}{\rulet}{E_1} ~\wedge~ 
     \dres{\tenv}{\rulet}{E_2} \quad\Rightarrow\quad 
     \fctx{\etrans{\tenv}}{E_1}{E_2}{\ttrans[\rulet]} \]
\end{corollary}
Here the judgment $\fctx{\tenv}{E_1}{E_2}{T}$
denotes \emph{contextual equivalence}, which captures that
$E_1$ and $E_2$ behave the same in any well-typed program context. We do not give the
formal definition here as it is not required to prove the corollary; we only
need to know that it is an equivalence relation, and, more spefically, that
it is reflexive, i.e., any well-typed System~F term is contextually equivalent
to itself.

Thirdly, on top of the immediate coherence of deterministic resolution, 
an additional stability property holds.  
\begin{lemma}[Stability of Resolution]
Resolution is stable under type substitution.
\[\forall \tenv,\alpha,\tenv',\sigma,\rulet, E: \enskip\
\dres{\tenv,\alpha,\tenv'}{\rulet}{E} \enskip\wedge\enskip \tenv \vdash \sigma
\enskip\Rightarrow\enskip 
\dres{\tenv,\tenv'[\sigma/\alpha]}{\rulet[\sigma/\alpha]}{E[\ttrans[\sigma]/\alpha]} \]
\end{lemma}%
This is a key lemma to establish that static reduction of type application 
preserves typing and elaboration.
\begin{lemma}[Stability of Typing under Type Application Reduction]
Static reduction of type application preserves typing.
\[\forall \tenv,\alpha,\sigma,\rulet, E: \enskip\
\wte{\tenv}{(\Lambda \alpha.e)\,\suty}{\rulet}{(\Lambda \alpha.E)\,\ttrans[\suty]}
\enskip\Rightarrow\enskip 
\wte{\tenv}{e[\suty/\alpha]}{\rulet}{E[\ttrans[\suty]/\alpha]}
   \]
\end{lemma}%
This, together with another property of System~F's contextual equivalence:
\[ \forall \tenv, \alpha, E, T, T': \enskip\
     \fwte{\tenv}{(\Lambda \alpha.E)\,T}{T'} \enskip\Rightarrow\enskip 
     \fctx{\tenv}{(\Lambda \alpha.E)\,T}{E[T/\alpha]}{T'}\]
allows us to conclude the correctness of static reduction of type application.
\begin{theorem}[Correctness of Type Application Reduction]
If a type application and its reduced form elaborate to two System~F terms,
those terms are contextually equivalent.
\begin{equation*}
\forall \tenv, \alpha, e, \suty, E1, E2 : \enskip\\
\wte{\tenv}{(\Lambda \alpha.e)\,\suty}{\rulet}{E_1} ~\wedge~
\wte{\tenv}{e[\suty/\alpha]}{\rulet}{E_2}
\enskip\Rightarrow\enskip 
\fctx{\etrans{\tenv}}{E_1}{E_2}{\ttrans[\rulet]}
\end{equation*}
\end{theorem}
